Determining the alternatives for scalar implicature

### 2016-02-01 Paper/Paper.Rmd
bp pass sunday afternoon

But what are the alternatives that should be considered in the implicature computation more generally? Under classic accounts, listeners consider only those words whose meanings entail the actual message [@horn1972], and these alternatives enter into conventionalized or semi-conventionalized scales [@levinson2000]. For example, because *all* entails *some*, and hence is a "stronger" meaning, *all* should be considered as an alternative to *some* in implicatures. Similar scales exist for non-quantifier scales, e.g. *loved* entails *liked* (and hence "I liked the movie" implicates that I didn't love it).

@@

$$p_{S_1}(u \mid m) \propto p_{L_0} (m \mid u)$$

@@

$$p_{S_1}(u \mid m) \propto e^{-\alpha[-log(p_{L_0}(m \mid u)) - C(u)]}$$

@@

The two-alternatives condition included only the scalar pairs from @vantiel2014. The four-alternatives condition included the two scalar items plus the top two-alternatives generated for each scalar family by participants in Experiment 2. The five-alternatives condition included the four previous items plus one more neutral item chosen from those alternatives generated in Experiment 2. See Figure \ref{fig:allScalesTable} for the complete list of alternatives used in each model.

@@

```{r exp1Plots, fig.env = "figure*", fig.pos = "t", fig.width=6.5, fig.height=4, fig.align='center', set.cap.width=T, num.cols.cap=2, fig.cap = "Literal listener judgments from Experiments 1a,b,c. Proportion of participants indicating compatibility (answering 'yes') is shown on the vertical axis, with the horizontal axis showing number of stars on which the utterance was judged. Rows are grouped by scale and items within rows are ordered by valence. Colors indicate the specific experiment (1a,b,c) with experiments including different numbers of items."}

@@

# glmer WITHOUT experiment and with random effect for subjects
glm1 <- glmer(judgment ~ scale + stars + (1 | word) + (1 | workerid),
              family=  "binomial", data = allLiteralListener)
# glmer WITH experiment and with random effect for subjects
glm2 <- glmer(judgment ~ exp + scale +  stars + (1 | word) +  (1 | workerid),
              family = "binomial", data = allLiteralListener)
sumGlm2 <- summary(glm2)
anovaResults <- anova(glm1, glm2)
```
To examine this last issue of consistency across experiments (our secondary hypothesis) we ran a mixed effects model regressing compatibility judgments on scale, number of stars and experimental condition, with subject and word level random effects, which was the maximal structure that converged. Results indicate no significant differences between Experiment 1a and 1c, $b =$`r round(sumGlm2$coefficients[3], 2)`, $Z =$ `r round(sumGlm2$coefficients[19], 2)`, $p =$ `r round(sumGlm2$coefficients[27], 2)` or between 1b and 1c, $b =$ `r round(sumGlm2$coefficients[2], 2)`, $Z =$ `r round(sumGlm2$coefficients[18], 2)` $p =$ `r round(sumGlm2$coefficients[26], 2)` and the addition of condition as a predictor did not significantly improve model fit when compared to a model without the experiment condition variable using ANOVA $\chi^2($ `r anovaResults$"Chi Df"[2]` $) =$ `r round(anovaResults$Chisq[2], 2)`, $p =$ `r round(anovaResults$"Pr(>Chisq)"[2], 2)`. These findings suggest that literal semantic intuitions are not effected by the set of alternatives queried in our paradigm. This finding is important because literal semantic data generated in these three conditions are used to simulate the effects of different alternative sets on implicature generation in our model.

@@

To examine the effect different alternative sets have on implicature, we needed a way of selecting plausible alternatives. We addressed this issue by adopting a modified cloze task to elicit empirical alternatives for the entailment items we used in Experiment 1a and 3a. This design was inspired by Experiment 3 of @vantiel2014.

@@

# plotData <- head(plotData[order(plotData$n, decreasing = TRUE),], n = 24)
ggplot(data = plotData, aes(x = reorder(alt, -n), y = n)) +
  geom_bar(stat="identity") + 
  ylim(0, 52) +
  labs(x = "alternatives", y = "counts") +
  theme_bw() +
  theme(strip.text.x = element_text(size = 8),
        legend.position = "bottom",
        legend.margin = unit(.02,"in"),
        legend.key = element_blank(), 
        legend.key.size = unit(.1, "in"),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5, size = 6),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))
```

```{r exp2Plots, fig.env = "figure*", fig.pos = "t", fig.width=6.5, fig.height=2.5, fig.align='center', set.cap.width=T, num.cols.cap=2, fig.cap = "Pragmatic listener judgements for scalar items. Proportion of participants generating a star rating is shown on the vertical axis, with the horizontal axis showing number of stars on which the utterance was judged. Line type denotes condition, and colors indicate the particular scalar items. Each panel shows one scalar pair, with only entailment items (two-alternatives condition) shown here for simplicity."}
# Get zero's in data frame
baseDf <- data.frame(degree = rep(c("hi", "low", "hi1", "hi2", "low1", "low2"), 25),
                     scale = c(rep("good_excellent", 6 * 5),
                               rep("liked_loved", 6 * 5),
                               rep("memorable_unforgettable", 6 * 5),
                               rep("palatable_delicious", 6 * 5),
                               rep("some_all", 6 * 5)),
                     stars = rep(1:5, 30),
                     combine = rep(TRUE, 150))
listeners <- bind_rows(listenerE6 %>% mutate(expt = "e6"), 
                      listenerE11 %>% mutate(expt = "e11"))
listeners <- left_join(baseDf, listeners) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p),
         expt = ifelse(degree == "hi" | degree == "low", "e6", "e11"))
listeners$word <- as.factor(with(listeners, mapply(lookupScalar, degree, scale, expt)))
listeners$plotThis <- with(listeners, ifelse(scale == "some_all" &
                               (degree == "hi" | degree == "low" | degree == "hi1" | degree == "low1"),
                               TRUE, ifelse(scale != "some_all" & (
                                 degree == "hi" | degree == "low" | degree == "hi1" | degree == "hi2"),
                                 TRUE, FALSE)))
listeners$`Scalar type` <- with(listeners, ifelse(degree == "hi" | degree == "hi1", "stronger", "weaker"))
listeners$Condition <- with(listeners, ifelse(expt == "e6", "two alts", "four alts"))

@@

Figure \ref{fig:exp2_altsPlot_likedLoved} shows an example alternative set for the scalar items *liked* and *loved* (combined). We expanded our alternative set from only entailment items (Experiments 1a, 3a) to four alternatives (Experiments 1b, 3b) by including the top two alternatives elicited for each scale in Experiment 2. For example, the top two alternatives for the *liked*/*loved* scale were *hated* and *disliked* and were included in four-alternatives model runs. Similarly, neutral items (e.g. *okay* in the *good*/*excellent* scalar family) were chosen from among the set of alternatives elicited in Experiment 2. Alternative set distributions for the other scalar families (e.g. *good*/*excellent*, *memorable*/*unforgettable*) were similarly long-tailed. See Figure \ref{fig:allScalesTable} for the complete list of alternatives we used from Experiment 2.

@@

Experiment 3 was conducted to measure pragmatic judgments---$p_{L_1}(m \mid u)$ in Equation $(1)$. As in Experiment 1, we include several conditions to test inferences in the presence of different alternative sets. In the two-alternatives condition (Experiment 3a), participants made judgments for items included in the entailment scales. In the four-alternatives condition (Experiment 3b), participants made judgments for the entailment items and also the top two-alternatives elicited for each scale in Experiment 2. Including two conditions with differing alternatives allowed us to rule out the potential effects of having a larger set of alternatives during the pragmatic judgment elicitation and also provided two sets of human judgments to compare with model predictions.\footnote{Note that alternatives in the four-alternatives condition were chosen on the basis of Experiment 2, which was run chronologically after the two-alternatives condition; both pragmatic listener experiments are grouped together for simplicity in reporting.}

@@

Figure \ref{fig:exp2Plots} plots pragmatic listener judgments distributions for "weak" / "strong" scalar pairs (e.g. *good*/*excellent*). We include only the original entailment pairs in this figure for simplicity. Several trends are visible. First, in each scale participants generated implicatures. They were substantially  less likely to assign high star ratings to weaker scalar terms, despite the literal semantic compatibility of those terms with those states shown in Experiment 1. Second, the size of the difference between strong and weak scalar items varied considerably across scales, consistent with previous work [@vantiel2014].

@@

Using literal listener data from Experiment 1, we conducted a set of simulations with the RSA model. Each simulation kept the model constant, fitting the choice parameter $\alpha$ as a free parameter, but used a set of alternatives to specify the scale over which predictions were computed. We considered four different alternative sets, with empirical measurements corresponding to those shown in Figure \ref{fig:allScalesTable}: 1) the two-alternatives in the classic entailment scales, 2) those two alternatives with the addition of a generic negative alternative \footnote{literal semantics for the generic negative alternative were not generated empirically, but serve as a baseline "none" style semantics in which the scalar item is only compatible with 1 star}, 3) the expanded set of four alternatives tested in Experiment 1b, and 4) the expanded set of five alternatives tested in Experiment 1c.

>>>>

But what are the alternatives that should be considered in the implicature computation more generally? Under classic accounts, listeners consider only those words whose meanings entail the actual message [@horn1972], and these alternatives enter into conventionalized or semi-conventionalized scales [@levinson2000]. For example, because *all* entails *some*, and hence is a "stronger" meaning, *all* should be considered as an alternative to *some* in implicatures. Similar scales exist for non-quantifier scales, e.g. *loved* entails *liked* (and hence "I liked the movie" implicates I didn't love it).

@@

$$p_{S_1}(u \mid m) \propto p_{L_0} (m \mid u) \tag{2}$$

@@

$$p_{S_1}(u \mid m) \propto e^{-\alpha[-log(p_{L_0}(m \mid u)) - C(u)]} \tag{3}$$

@@

The two-alternatives condition included only the scalar pairs from @vantiel2014. The four-alternatives condition included the two scalar items plus the top two alternatives generated for each scalar family by participants in Experiment 2. The five-alternatives condition included the four previous items plus one more neutral item chosen from alternatives generated in Experiment 2. See Figure \ref{fig:allScalesTable} for the complete list of alternatives used in each condition.

@@

```{r exp1Plots, fig.env = "figure*", fig.pos = "t", fig.width=6.5, fig.height=4, fig.align='center', set.cap.width=T, num.cols.cap=2, fig.cap = "Literal listener judgments from Experiment 1. Proportion of participants indicating compatibility (answering `yes') is shown on the vertical axis, with the horizontal axis showing number of stars on which the utterance was judged. Rows are grouped by scale and items within rows are ordered by valence. Colors indicate the specific condition (1a,b,c) with conditions including different numbers of items."}

@@

# glmer WITHOUT experiment and with random effect for subjects
glm1 <- glmer(judgment ~ scale + stars + (1 | word) + (1 | workerid),
              family=  "binomial", data = allLiteralListener)
# glmer WITH experiment and with random effect for subjects
glm2 <- glmer(judgment ~ exp + scale +  stars + (1 | word) +  (1 | workerid),
              family = "binomial", data = allLiteralListener)
sumGlm2 <- summary(glm2)
anovaResults <- anova(glm1, glm2)
```
To examine this last issue of consistency across experiments (our secondary hypothesis) we fit a mixed effects model, regressing compatibility judgment on scale, number of stars and experimental condition, with subject- and word-level random effects, which was the maximal structure that converged. Results indicate no significant differences between Experiment 1a and 1c, $b =$`r round(sumGlm2$coefficients[3], 2)`, $Z =$ `r round(sumGlm2$coefficients[19], 2)`, $p =$ `r round(sumGlm2$coefficients[27], 2)` or between 1b and 1c, $b =$ `r round(sumGlm2$coefficients[2], 2)`, $Z =$ `r round(sumGlm2$coefficients[18], 2)` $p =$ `r round(sumGlm2$coefficients[26], 2)` and the addition of condition as a predictor did not significantly improve model fit when compared to a model without the experiment condition variable using ANOVA $\chi^2($ `r anovaResults$"Chi Df"[2]` $) =$ `r round(anovaResults$Chisq[2], 2)`, $p =$ `r round(anovaResults$"Pr(>Chisq)"[2], 2)`. These findings suggest that literal semantic intuitions are not effected by the set of alternatives queried in our paradigm. This finding is important because literal semantic data generated in these three conditions are used to simulate the effects of different alternative sets on implicature generation in our model.

@@

To examine the effect different alternative sets have on implicature, we needed a way of expanding alternative sets beyond entailment pairs. We addressed this issue by adopting a modified cloze task to elicit alternatives empirically. This design was inspired by Experiment 3 of @vantiel2014.

@@

# plotData <- head(plotData[order(plotData$n, decreasing = TRUE),], n = 24)
ggplot(data = plotData, aes(x = reorder(alt, -n), y = n)) +
  geom_bar(stat="identity") + 
  ylim(0, 52) +
  labs(x = "alternatives", y = "counts") +
  theme_bw() +
  theme(strip.text.x = element_text(size = 8),
        legend.position = "bottom",
        # legend.margin = unit(.02,"in"),
        legend.key = element_blank(), 
        # legend.key.size = unit(.1, "in"),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0, size = 6),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))
```

```{r exp2Plots, fig.env = "figure*", fig.pos = "t", fig.width=6.5, fig.height=2.5, fig.align='center', set.cap.width=T, num.cols.cap=2, fig.cap = "Pragmatic listener judgements for scalar items. Proportion of participants generating a star rating is shown on the vertical axis, with the horizontal axis showing number of stars on which the utterance was judged. Line type denotes condition, and colors indicate the particular scalar items. Each panel shows one scalar pair, with only entailment items (two-alternatives condition) shown here for simplicity in reporting."}
# Get zero's in data frame
baseDf <- data.frame(degree = rep(c("hi", "low", "hi1", "hi2", "low1", "low2"), 25),
                     scale = c(rep("good_excellent", 6 * 5),
                               rep("liked_loved", 6 * 5),
                               rep("memorable_unforgettable", 6 * 5),
                               rep("palatable_delicious", 6 * 5),
                               rep("some_all", 6 * 5)),
                     stars = rep(1:5, 30),
                     combine = rep(TRUE, 150))
listeners <- bind_rows(listenerE6 %>% mutate(expt = "e6"), 
                      listenerE11 %>% mutate(expt = "e11"))
listeners <- left_join(baseDf, listeners) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p),
         expt = ifelse(degree == "hi" | degree == "low", "e6", "e11"))
listeners$word <- as.factor(with(listeners, mapply(lookupScalar, degree, scale, expt)))
listeners$plotThis <- with(listeners, ifelse(scale == "some_all" &
                               (degree == "hi" | degree == "low" | degree == "hi1" | degree == "low1"),
                               TRUE, ifelse(scale != "some_all" & (
                                 degree == "hi" | degree == "low" | degree == "hi1" | degree == "hi2"),
                                 TRUE, FALSE)))
listeners$`Scalar type` <- with(listeners, ifelse(degree == "hi" | degree == "hi1", "stronger", "weaker"))
listeners$Condition <- with(listeners, ifelse(expt == "e6", "two alts", "four alts"))

@@

Figure \ref{fig:exp2_altsPlot_likedLoved} shows an example alternative set for the scalar items *liked* and *loved* (combined). We expanded our alternative set from only entailment items (items used in Experiments 1a, 3a) to four alternatives (Experiments 1b, 3b) by including the top two alternatives elicited for each scale in Experiment 2. For example, the top two alternatives for the *liked*/*loved* scale were *hated* and *disliked* and were included in four-alternatives model. Similarly, neutral items (e.g. *okay* in the *good*/*excellent* scale) were chosen from among the set of alternatives elicited in Experiment 2. See Figure \ref{fig:allScalesTable} for the complete list of alternatives we used from Experiment 2.

@@

Experiment 3 was conducted to measure pragmatic judgments---$p_{L_1}(m \mid u)$ in Equation $(1)$. As in Experiment 1, we include several conditions to test inferences in the presence of different alternative sets. In the two-alternatives condition (Experiment 3a), participants made judgments for items included in the entailment scales. In the four-alternatives condition (Experiment 3b), participants made judgments for the entailment items and also the top two-alternatives elicited for each scale in Experiment 2. Including two conditions with differing alternatives allowed us to rule out the potential effects of having a larger set of alternatives during the pragmatic judgment elicitation and also provided two sets of human judgments to compare with model predictions.\footnote{Alternatives in the four-alternatives condition were chosen on the basis of Experiment 2, which was run chronologically after the two-alternatives condition; both pragmatic listener experiments are grouped together for simplicity in reporting.}

@@

Figure \ref{fig:exp2Plots} plots pragmatic listener judgments distributions for "weak" / "strong" scalar pairs (e.g. *good*/*excellent*). Several trends are visible. First, in each scale participants generated implicatures. They were substantially  less likely to assign high star ratings to weaker scalar terms, despite the literal semantic compatibility of those terms with those states shown in Experiment 1. Second, the size of the difference between strong and weak scalar items varied considerably across scales, consistent with previous work [@vantiel2014].

@@

Using literal listener data from Experiment 1, we conducted a set of simulations with the RSA model. Each simulation kept the model constant, fitting the choice parameter $\alpha$ as a free parameter, but used a set of alternatives to specify the scale over which predictions were computed. We considered four different alternative sets, with empirical measurements corresponding to those shown in Figure \ref{fig:allScalesTable}: 1) the two-alternatives in the classic entailment scales, 2) those two alternatives with the addition of a generic negative alternative\footnote{Literal semantics for the generic negative alternative were not generated empirically, but serve as a baseline ``none" style semantics in which the scalar item is only compatible with 1 star.}, 3) the expanded set of four alternatives tested in Experiment 1b, and 4) the expanded set of five alternatives tested in Experiment 1c.
